Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2015-03-09 pituba/article/article_spe.tex
artigo

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium, i.e., $\vec{u}_\alpha = 0$. This implies that pressure at some time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.

@@

It will be shown that the pressure field can be further simplified by neglecting the variation of density with depth. For instance, the average oil density can be approximated to the local density at the datum level, $\rho_o (t,z) \approx \rho_o (t, z_{datum})$, what results in the following simple expression for the pressure field:
\begin{equation}
p_\alpha (t,z) = p_\alpha \left(t, z_{datum}\right)+ \rho_\alpha (t, z_{datum}) g \left(z-z_{datum}\right) \, .
\end{equation}

@@

\subsection{Solution algorithm}
Several methods exist to find roots of nonlinear continuous functions \citep{hamming1987numerical}. One possible choice is Newton's method, which linearizes the function using either its gradient or a secant line. The secant line approach introduces a numerical smoothness to the gradient near continuous but non-differentiable points, for what it is the approach used here to deal with the bubble-point pressure.

An open source implementation of the secant method is available in SciPy library \citep{scipy}. This work used the library version $0.12$.

>>>>

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium: $\vec{u}_\alpha = 0$. This implies that pressure at some time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.

@@

It will be shown that the pressure field can be further simplified by neglecting the variation of density with depth, what allows us to express the pressure field in the following simple expression:
\begin{equation}
p_\alpha (t,z) = p_\alpha \left(t, z_{datum}\right)+ \rho_\alpha (t, z_{datum}) g \left(z-z_{datum}\right) \, .
\end{equation}

@@

\subsection{Solution algorithm}
Several methods exist to find roots of nonlinear continuous functions \citep{hamming1987numerical}. One possible choice is Newton's method with a switch from the tangent to the secant line near the non-differentiable bubble-point pressure. Both methods are available in the open source SciPy library \citep{scipy}. This work used the library version $0.12$.
